---
title: "Misc. data viz"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

This page to host some misc. data visualization and notes for phylogenetics.

## Haemoprotozoa prevalence

**Import data**
```{r}
library(readr)
haem <- read_csv("data/haemoprotozoa.csv")
```

### UpSetR plot

Creating [UpSetR](https://github.com/hms-dbmi/UpSetR) and the add on [ComplexUpset](https://krassowski.github.io/complex-upset/) plot to show co-infections, 
Extra examples [here](https://jokergoo.github.io/ComplexHeatmap-reference/book/upset-plot.html)

**Load libraries**

```{r libs, include=FALSE}
library(UpSetR)
library(tidyverse)
```

**Load data**

Must be in a dataframe format
```{r}
# convert to dataframe
haemdf <- as.data.frame(haem)
str(haemdf)
# define columns with species info
haem_sp = colnames(haemdf)[3:9]
haem_sp
```

Simple UpSetR plot
```{r}
upsetR1 = upset(haemdf, sets = c("Hepatozoon sp.", "Babesia lohae-like", "Theileria cf. peramelis", "Trypanosoma lewisi-like", "Trypanosoma noyesi", "Trypanosoma gilletti", "Trypanosoma cyclops-like"), sets.bar.color = "#7a255d")
upsetR1
```


More detailed plots using [ComplexUpset](https://krassowski.github.io/complex-upset/)
```{r}
library(ComplexUpset)

upsetR3 = upset(
    haemdf,
    haem_sp, width_ratio=0.1,
    base_annotations=list(
        'Intersection size'=intersection_size(
            counts=TRUE,
            mapping=aes(fill=species_code)
        ))
)
upsetR3


upsetR4 = upset(
    haemdf, haem_sp,
    set_sizes=(
        upset_set_size(
            geom=geom_bar(
                aes(fill=species_code, x=group)
            ),
            position='right'
        )
    ),
    # moves legends over the set sizes
    guides='over'
)

upsetR4
```

## Phylogenetics

All phylogenetics created in related manuscript was done using IQTEE.

My detailed notes on using IQTREE server for generated phylogenies is [here](http://siobhonlegan.com/research_site/bioinfo/phylogenetics/iqtree.html).

Trees visualised using [FigTree](http://tree.bio.ed.ac.uk/software/figtree/) and annotated in [inkscape](https://inkscape.org/).

