---
title: "Trypanosoma phyloseq"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

Visualization and analysis of microbiome data.

> Note this was run using R version 4.0.3 and RStudo version 1.4. See [R info](#rinfo) for full details of R session.


## 0. Load libraries

**Install libraries if required**

Only need to run this code once.

```
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
# phyloseq
source('http://bioconductor.org/biocLite.R')
biocLite('phyloseq')
#tidyverse
install. packages("tidyverse")
#ampvis2
install.packages("remotes")
remotes::install_github("MadsAlbertsen/ampvis2")
#ampvis2extras
install.packages("BiocManager")
BiocManager::install("kasperskytte/ampvis2extras")
#ggpubr
install.packages("ggpubr")
#agricolae
install.packages("agricolae")
install.packages("remotes")
remotes::install_github("DanielSprockett/reltools")
devtools::install_github('jsilve24/philr')
#decontam
BiocManager::install("decontam")
library(decontam)
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("Biostrings")
```

**Load libraries**

```{r}
library(qiime2R)
library(phyloseq)
library(tidyverse)
library(ampvis2)
library(ampvis2extras)
library(ggpubr)
library(agricolae)
library(plotly)
library(viridis)
library(cowplot)
library(MicrobeR)
library(microbiome)
library(reshape)
library(decontam)
library(data.table)
library(ape)
library(DESeq2)
library(vegan)
library(microbiomeutilities)
library(knitr)
library(tibble)
library(dplyr)
library(patchwork)
library(Biostrings)
```
## 1. Import data

Generate phyloseq object from spreadsheets.

Import ASV/OTU count data
```{r}
count_data <- read_csv("data/tryp-phyloseq/count_data.csv")
# use first column as label for rows
count_data_lab = column_to_rownames(count_data, var = "#Zotu ID")
# Make matrix
otumat <- as.matrix(count_data_lab)
```
Import taxonomy data
```{r}
taxonomy <- read_csv("data/tryp-phyloseq/taxonomy.csv", 
    col_types = cols(Accession_description = col_skip(), 
        `Accession no.` = col_skip(), evalue = col_skip(), 
        `per. Ident` = col_skip(), taxid = col_skip()))
# use first column as label for rows
taxonomy_lab = column_to_rownames(taxonomy, var = "#Zotu ID")
taxmat <- as.matrix(taxonomy_lab)
```

Check the class of the otumat and taxmat objects, they MUST be in matrix format. Then we can great a phyloseq object called physeq from the otu and taxonomy tables and check the sample names.
```{r}
class(otumat)
class(taxmat)
OTU = otu_table(otumat, taxa_are_rows = TRUE)
TAX = tax_table(taxmat)
physeq = phyloseq(OTU, TAX)
physeq
sample_names(physeq)
```

**Add metadata and sequence data**

Add sequences to phyloseq object
```{r}
# read sequence file
rep.seqs <- Biostrings::readDNAStringSet("data/tryp-phyloseq/unoise_zotus.fasta", format = "fasta")
```

Add metadata, importing gDNAID as factor to be able to merge later on
```{r}
metadata <- read_csv("data/tryp-phyloseq/sampledata.csv")
metadata_lab = column_to_rownames(metadata, var = "SampleID")
sampledata = sample_data(data.frame(metadata_lab))
sampledata

```

**Create final phyloseq object**

Now you can merge your data to create a final phyloseq object
```{r eval=FALSE}
ps_raw_tryp = merge_phyloseq(physeq, sampledata, rep.seqs)
```

Remove samples with NA values or not part of final data set,
```{r}
ps_raw_tryp <- subset_samples(ps_raw_tryp, !SampleType=="SampleEcol")
```

An easy way to view the tables is using Nice Tables
```{r}
Nice.Table(ps_raw_tryp@sam_data)
```
```{r}
Nice.Table(ps_raw_tryp@tax_table)
```


## 2. QC plots

R package `decontam` to assess contaminating OTUs, [tutorial](https://bioconductor.org/packages/release/bioc/vignettes/decontam/inst/doc/decontam_intro.html).

The CRAN version only works on R version <4.
To install for R versions >4 install from [bioconductor](https://bioconductor.org/packages/release/bioc/html/decontam.html) using the following

Make plot of library size of Samples vs Controls
```{r}
df <- as.data.frame(sample_data(ps_raw_tryp)) # Put sample_data into a ggplot-friendly data.frame
df$LibrarySize <- sample_sums(ps_raw_tryp)
df <- df[order(df$LibrarySize),]
df$Index <- seq(nrow(df))
libQC <- ggplot(data=df, aes(x=Index, y=LibrarySize, color=SampleType)) + geom_point() + theme_bw() + scale_colour_brewer(palette = "Set1")
ggsave("libQC.pdf", plot = libQC, path = "output/plots/trypNGS", width = 10, height = 10, units = "cm")
```

Make html plot with `plotly`
```{r}
libQCplotly <- ggplotly(libQC)
htmlwidgets::saveWidget(libQCplotly, "output/plots/libQCplotly.html")
```


Make distribution plot of reads using [microbiomeutilities](https://microsud.github.io/microbiomeutilities/articles/microbiomeutilities.html#distribution-of-reads-1)
```{r}
distrib <- plot_read_distribution(ps_raw_tryp, groups = "SampleCategory", 
                            plot.type = "density") + xlab("Reads per sample") + ylab("Density")
distrib <- distrib + geom_density(alpha = 0.5, fill = "grey") + theme(axis.text.x = element_text(angle = 45, size=10, vjust = 1))
ggsave("distrib1.pdf", plot = distrib, path = "output/plots/trypNGS", width = 15, height = 10, units = "cm")
```

## 3. Subset and Filter

Filter out negative controls and after identifying threshold for positive reads, transform otu count data 
```{r}
ps_dec_tryp <- subset_samples(ps_raw_tryp, !SampleType=="ControlNeg")
ps_dec_tryp@otu_table [, 1:492][ps_dec_tryp@otu_table [, 1:492] < 100] <- 0
```

Save R data for phyloseq object - saving "raw data" (`ps_raw_tryp`) and "decontaminated data"  (`ps_dec_tryp`)
```{r}
save(ps_raw_tryp, file = "../data/Rdata/ps_raw_tryp.RData")
save(ps_dec_tryp, file = "../data/Rdata/ps_dec_tryp.RData")
```

To load raw and decon data quickly from `.RData` format.
```{r}
load("../data/Rdata/ps_raw_tryp.RData")
load("../data/Rdata/ps_dec_tryp.RData")
```

### 3.1. For phyloseq object

Subset phyloseq object based on sample types
```{r}
# Exclude controls
ps_tryp_samp = subset_samples(ps_dec_tryp, !SampleType=="ControlPos")
# Blood samples only
ps_tryp_bl = subset_samples(ps_tryp_samp, SampleCategory=="Blood")
# Tissue samples only
ps_tryp_tis = subset_samples(ps_tryp_samp, SampleCategory=="Tissue")
# Tick samples only
ps_tryp_tick = subset_samples(ps_tryp_samp, SampleCategory=="Tick")
```


Subset phyloseq object based on host species
```{r}
# Black rat
ps_BR = subset_samples(ps_tryp_samp, species=="Black rat")
# Brush tail possum
ps_BTP = subset_samples(ps_tryp_samp, species=="Brush tail possum")
# Chuditch
ps_chud = subset_samples(ps_tryp_samp, species=="Chuditch")
# Long-nosed bandicoot
ps_LNB = subset_samples(ps_tryp_samp, species=="Long-nosed bandicoot")
```

### 3.2. For ampvis2 object

**Make ampvis2 object for analysis**

```{r}
#require the devtools package to source gist
if(!require("devtools"))
  install.packages("devtools")
#source the phyloseq_to_ampvis2() function from the gist
devtools::source_gist("8d0ca4206a66be7ff6d76fc4ab8e66c6")

#convert
tryp_amp <- phyloseq_to_ampvis2(ps_dec_tryp)
```

Subset ampvis2 object based on sample category
```{r}
#remove controls
amp_samp <- amp_subset_samples(tryp_amp, 
                                 !SampleType %in% c("ControlPos"),
                                 RemoveAbsents = TRUE)
#blood samples
amp_bl <- amp_subset_samples(amp_samp, 
                                 SampleCategory %in% c("Blood"),
                                 RemoveAbsents = TRUE)
#tissue samples
am_tis <- amp_subset_samples(amp_samp, 
                                 SampleCategory %in% c("Tissue"),
                                 RemoveAbsents = TRUE)
#tick samples
amp_tick <- amp_subset_samples(amp_samp, 
                                 SampleCategory %in% c("Tick"),
                                 RemoveAbsents = TRUE)
```

## 4. Heatmap

### 4.1. ampvis2

Make heat map using [ampvis2](https://madsalbertsen.github.io/ampvis2/reference/amp_heatmap.html)

First filter taxa 
```{r}
# order Trypanosomatida
tax_vector1 <- c(
  "Trypanosomatida"
)

amp_samp_otry <- amp_subset_taxa(amp_samp,
  tax_vector = tax_vector1)

# Trypanosoma species of interest
tax_vector2 <- c(
  "Trypanosoma gilletti",
  "Trypanosoma sp. (cyclops-like)",
  "Trypanosoma vegrandis",
  "Trypanosoma noyesi",
  "Trypanosoma sp. (lewisi-like)",
  "Lotmaria passim"
)

amp_samp_targettryp <- amp_subset_taxa(amp_samp,
  tax_vector = tax_vector2)
```


**Relative abundance**

Heatmap of subsetted data using relative abundance
```{r}
# Relative abundance order level
heatmap_rel1 <- amp_heatmap(amp_samp_otry,
            facet_by = "SampleCategory",
            group_by = "species",
            tax_aggregate = "Species",
            tax_show = 10,
            normalise = TRUE,
            plot_values = FALSE,
            plot_values_size = 3,
            round = 0, color_vector = c("white", "#e5ba52", "#ab7ca3", "#9d02d7", "#0030bf"), plot_colorscale = "log10")  +
  theme(axis.text.x = element_text(angle = 45, size=10, vjust = 1),
        axis.text.y = element_text(size=10),
        legend.position="right")

# Relative abundance trypanosoma species of interest
heatmap_rel2 <- amp_heatmap(amp_samp_targettryp,
            facet_by = "SampleCategory",
            group_by = "species",
            tax_aggregate = "Species",
            tax_show = 10,
            normalise = TRUE,
            plot_values = FALSE,
            plot_values_size = 3,
            round = 0, color_vector = c("white", "#e5ba52", "#ab7ca3", "#9d02d7", "#0030bf"), plot_colorscale = "log10")  +
  theme(axis.text.x = element_text(angle = 45, size=10, vjust = 1),
        axis.text.y = element_text(size=10),
        legend.position="right")

```
Save PDF of plots
```{r}
ggsave("heatmap_rel1.pdf", plot = heatmap_rel1, path = "output/plots/trypNGS", width = 25, height = 15, units = "cm")
ggsave("heatmap_rel2.pdf", plot = heatmap_rel2, path = "output/plots/trypNGS", width = 25, height = 15, units = "cm")
```

**Count of sequences**
```{r}
heatmap_count <- amp_heatmap(amp_samp,
            facet_by = "SampleCategory",
            group_by = "species",
            tax_aggregate = "Family",
            tax_show = 40,
            normalise = FALSE,
            plot_values = FALSE,
            plot_values_size = 3,
            round = 0, color_vector = c("white", "#e5ba52", "#ab7ca3", "#9d02d7", "#0030bf"), plot_colorscale = "log10") + theme(axis.text.x = element_text(angle = 45, size=10, vjust = 1), axis.text.y = element_text(size=10), legend.position="right")
```

### 4.2. microbiomeutilities

Heatmap using [microbiomeutilities](https://microsud.github.io/microbiomeutilities/articles/microbiomeutilities.html#heatmaps)

Subset taxa
```{r}
ps_tryp_otry = subset_taxa(ps_tryp_samp, Order=="Trypanosomatida")
ps_tryp_subtry = subset_taxa(ps_tryp_samp, Species=="Trypanosoma gilletti" | Species=="Trypanosoma sp. (cyclops-like)" | Species =="Trypanosoma vegrandis" | Species=="Trypanosoma noyesi" | Species=="Trypanosoma sp. (lewisi-like)" | Species =="Lotmaria passim")
```

```{r}

# create a gradient color palette for abundance
#grad_ab <- colorRampPalette(c("#faf3dd","#f7d486" ,"#5e6472"))
color_vector = colorRampPalette(c("#faf3dd", "#e5ba52", "#ab7ca3", "#9d02d7", "#0030bf"))
grad_ab_pal <- color_vector(10)

# create a color palette for varaibles of interest

meta_colors = list(c("Blood" = "#7a255d", "Tick" = "#9fd0cb", "Tissue" = "#7566ff"), 
                  c("Brush tail possum" = "#440154FF", "Black rat" = "#482878FF", "Swamp rat"="#3E4A89FF", "Long-nosed bandicoot" ="#31688EFF", "Bush rat"="#26828EFF", "Brown antechinus" = "#1F9E89FF", "Rabbit"="#35B779FF", "Chuditch"= "#6DCD59FF", "Quenda" = "#B4DE2CFF", "Deer" = "#FDE725FF" ))

# add labels for pheatmap to detect
names(meta_colors) <- c("SampleCategory", "species")

ph_heatmap <- plot_taxa_heatmap(ps_tryp_subtry,
                       subset.top = 50,
                       VariableA = c("SampleCategory","species"),
                       heatcolors = grad_ab_pal, #rev(brewer.pal(6, "RdPu")),
                       transformation = "log10",
                       cluster_rows = T,
                       cluster_cols = F,
                       show_colnames = F,
                       annotation_colors=meta_colors, fontsize = 8,)
```

Save PDF of plots
```{r}
ggsave("ph_heatmap.pdf", plot = ph_heatmap$plot, path = "output/plots/trypNGS", width = 25, height = 15, units = "cm")
```
